Setting the Stage for Understanding the Main Roles in Software Organizations

### 2011-09-28 soulgem.tex
Third draft complete.

There was no simple role classification in the organizations we studied. However, it appears that some patterns in role definitions would emerge with further studies. We can see the beginnings of such patterns in Table \ref{tableExpectations}, which shows the expectations placed on the people we interviewed (about half of the total number of organization members). The columns in the table represent our interviewees and the rows represent common or significant expectations. The reader should note that the list of expectations is necessarily incomplete for the general case. It is also somewhat subjective, as expectations were not always clearly articulated, and our participants did not agree on their expectation assignments in every case.

One pattern concerns the technical core of the organizations' products. Work on this area was reserved almost exclusively for developers and their technical and team leads. However, the leads' expectations were more spread out: technical leads focused more on architecture and design issues, whereas team leads emphasized coordination activities. There were also purely organizational roles that do not need any domain or technical expertise. But in both organizations some key people had more than one kind of expertise (technical, domain, and organizational), and these people seemed vital for the functioning of the organization as they transitioned between boundaries with ease. A representative example is a Gallium Field Specialist (the second column in the Support \& Field cluster), who straddled all three kinds of expertise. Several people in different areas pointed out his relevance; in the words of one informant, ``the company depends on [him] so much he should have armed guards behind him taking him to and from work.''

@@

\emph{Product Management at Gallium:} There are three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead. The latter documents the requirements of new functionality, he fulfills a ``historian'' expectation (sharing design rationale with developers and pointing out pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices. Of the two Product Managers, one has an inside-looking approach, whereas the other one has an outside-looking approach: he has a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimizes the organization. He also understands how the users would interact with the product and helps shape the user interface to bring it closer to the users' mental model. Other people in the organization also performed some Product Management activities unofficially: Customer Services Representatives and Field Specialists often persuade developers to focus on the features that cause them the most pain, for example. In sum, although the three employees above have similar goals, the expectations placed on them vary considerably, are not interchangeable, and are not their exclusive prerogative.

\emph{Team Leading at Gallium and Cobalt:} Every team in the sites we studied had a Team Lead, but this leadership role was tinged by the characteristics of the team and project in question. For instance, in Cobalt, team membership is fluid---every developer may be expected to work on any project, according to need---and some Team Leads performed coordination tasks, others had a dormant leadership position, waiting for new issues or concerns in their projects to arise, and yet others were more like ``first responders'' than coordinators, requesting assistance only if the work load for their projects became too much for themselves. In Gallium, there were three Team Leads at the time of our observations. They shared a few characteristics: they were expected to develop software, and they reported to their Director. But their roles were quite different. The Team Lead of the Alpha Team would spend about half of his time developing software, and in his other half, he meets with customer service representatives, project and product managers, performs estimation tasks, and analyzes the performance of his team. A second Team Lead is in charge of a product about to be discontinued, and that responsibility takes a small part of his time. He is also associated with the Beta Team, but since that team has an abundance of ``leads'' (it includes the two official Tech Leads of the organization), his coordination and reporting expectations are minimal. The third Team Lead has one other team member, a newcomer to the organization, and his leadership expectations consist mostly of mentoring and training rather than coordination and performance analysis.

\emph{Developers at Gallium and Cobalt:} The quintessential role in the software development literature was one of the hardest to pin down. A developer, of course, is expected to write production code, but what else? There is a multitude of answers to that question: some developers are expected to perform code reviews, some are expected to provide architectural input, some are expected to carry their own requirements analysis, some are expected to socialize, some are expected to interface with customers. None of the developers in our sample were expected to perform all of the above.

\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.\footnote{Note that, as opposed to the previous examples, this role has no clear label to refer to it.} But the stuff that falls through the cracks was different in both sites, and therefore, the expectations were different. In Cobalt, much of this referred to sheltering the organization from bureaucratic processes and constraints. In Gallium, it partly meant helping create and shape new and needed groups, and partly to perform activities that nobody else was tasked with doing, such as some kinds of documentation.

\emph{Project Management at Gallium:} These Project Managers are a counterexample to our findings, to some extent. Although there are distinctions between them (one has a Project Management Leader role, another creates templates and institutes new processes, and all, including a third and more junior Project Manager, implement them), the similarities in their roles seem to outweigh the differences. The main reason seems to be that these Project Managers bring a fairly specific and standardized way of doing things with them, in the form of a Project Management Body of Knowledge and certification. In other words, the institutionalization of their role is strong. Part of their strategy appears to be to define their role with sufficient precision, and to enforce their definition in the rest of the organization. No other role that we examined had the emphasis on role definition and consistency that the Gallium Project Managers had.

@@

We found that expectations are clustered in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors, and that these kinds of factors can interplay and provide feedback to each other.

\emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it helps shape role definitions. We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital. It quickly expanded and distributed its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them up with newcomers with the goal of taking care of those aspect of the business. Before this influx, the division of labour was more flexible and organic: most organization members did a bit of everything. After the influx, role definitions shifted towards a more stereotypical differentiation of roles and groups.

\emph{Historical factors:} Previous events in the history of the organization shape its current clustering of expectations, in the same way that alterations early in the life of other organic systems shape their later structure. Some developers that have been part of the organization for a long time are not just developers, they are also mentors, historians, and political players, despite what the organizational chart says about them. This affects the expectations that people place on them. During the Gallium downsizing, some people were demoted to keep the company afloat. At some point, one of them had been the head of the QA department (which was dispersed) and, later, the Program Manager. At the time of our observations he was a QA Analyst, yet he was still involved (and expected to be involved) in many strategic activities that do not match the institutional view of what a QA Analyst is supposed to do.

\emph{Structural factors:} A person's position in the organizational structure gives her a certain power and awareness that others may not have. Her structural position leads others to expect from her certain behaviours: she may be expected to play a bridging, directing, or even confrontational role. Many of the expectations placed on the head of the Operations group at Cobalt spawn from his structural position. He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from system administration concerns, and he is expected to control the gateway that determines whether a product is ready to be released or not---that is, to assess and have final say on the operational quality of the product. These expectations come from his position in the organization: it would be hard for someone without his position to satisfy them.

\emph{Personal factors:} A person's background, preferences, and personality shapes the expectations she receives from others. Perhaps her background is deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from her. Alternatively, his traits (curiosity, proactivity, a talent) may lead her to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors. For example, the Project Administrator at Cobalt was hired to perform clerical tasks, but his own preference led him to gradually take more and greater responsibilities. The organization learned to be dependent on such behaviour. When he temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement. Projects suffered quickly and significantly as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him. We should also note that at the time of our observations, the Project Administrator position was undergoing an official redefinition in order to match real expectations. In this way, personal factors were leading to historical and structural changes in the organization, so that future enactors of the role played by the Project Administrator would be affected by them.

@@

However, a person's position has limited power in defining their role. As we have seen above, the expectations placed on a person also come from historical and personal factors. An example out of several in our observations comes from one of the Gallium Product Managers. He was hired as a Product Manager to steer and coordinate the development of the company's products, but he had no experience in software development, thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization, he struggled to understand the level of control and direction he was expected to have. Over time, he learned to fill the void of control that he was expected to fill, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of his peers with the same or similar positions.

@@

Regarding organizational growth, there is a long sociological tradition exploring its role in specialization and division of labour, and our data are far from surprising. In fact, it should be clear from pure argument that organizational growth drives role evolution. The smallest organization is formed by a single person, and that person should perform all the activities required for the functioning of the organization, from sales to debugging. As the organization obtains more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members, it is rare for a software group to perform undifferentiated labour. Furthermore, growth carries a coordination weight: larger organizations need to assign some of their members the role of coordinating the work of others, or of interfacing between groups. That is, larger organizations have needs that are caused by their size---needs that smaller organizations do not have. With organizations that are similar to those we observed, some degree of specialization is well underway. We expect that, should we perform this study in a larger organization, we would discover further role specialization.

Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest. In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} the role differentiation in the organization. The silos that group members wanted to break were project-based silos. They discovered that assigning one person per project introduced vulnerabilities (when something went wrong in a project, only one person could intervene) and rework (as many projects consisted of similar modules that had to be developed independently). To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem. The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate and Subsystem Leads took care to help their teammates understand the particularities of their subsystems.

The other important factor driving role evolution in our findings was a mismatch between expectations and behaviour. The fact that expectations are placed does not mean that they are satisfied. There are many potential reasons for their lack of satisfaction: for instance, the person upon whom they are placed may not know or understand the expectation, it could be beyond her means, or it could be in conflict with other expectations. Therefore, an assignment of expectations that works in theory may not work in practice. Similarly, an assignment that does \emph{not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond their expectations. However, in both the under- and the over-performing case, with time, organization members grow to expect the typical performance of others, and a new expectation (and hence, a new role) takes hold.

Several examples from our data above illustrate this phenomenon and we offer one more here. The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being one of the main contacts for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction and next steps. In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization. As a result, he now plays an essential role in the development of the group's products and in the maintenance of its relations with its users, a role that the rest of the organization depends on for its good functioning.

@@

We expected to find that people enacting similar roles use similar communication media, and that they communicate with people enacting interfacing roles through other media. Although we found support for this claim, as we explain below, a biasing factor was the prevalence of face-to-face interactions in both of our field sites. People with similar or different roles communicate face-to-face regularly. Stand-up meetings, walks across the hall, and group meetings were commonplace.

Regarding other media, we found role-based differences in medium selection and use. The broadest medium---the one that had the greatest use among the most kinds of roles---was the issue tracking system. People with expectations regarding code production, quality assurance, customer service, field implementation, product management, and performance monitoring all used the issue tracking system to some extent. Its use was not uniform: some group members used it to emit messages, others were mostly listeners, and others were more interested in the aggregate than in the contents of particular electronic traces. This finding mirrors the observation of De Souza \emph{et al.}\ regarding issue tracking systems as boundary objects \cite{deSouza2003}.

The use of other media was more localized. In Gallium, IRC use was common among some developer teams: each team had one channel, and these were not frequented by any non-developers. Customer Support staff used Google Documents to keep track of some of its cases, and to provide an ongoing documentation of recent events and requests for support. Email and phone were more heavily used in the management side of the organization than in its technical core: developers and QA analysts rarely made or received phone calls, and they obtained most of the information they needed about their projects through the issue tracking system.

@@

We also explored role evolution and found that it is driven by organizational growth (which leads to specialization), and by mismatches between expectations and behaviour (which lead to an adjustment of expectations). Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites in terms of inter-role communications, with the qualification that issue-tracking systems were used differently by people enacting different roles.

@@

\emph{Research:} First, the fact that role definitions are not institutionally uniform and respond to local (historical, structural, personal) factors should help clarify common misunderstandings and assumptions made in research. A stereotypical developer, team lead, or tester, is rarely found in practice. Instead, thinking in terms of tasks and the people who are expected to perform them will help us reattach our studies and findings to software development in the real world.

Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual). This in turn should lead to better role definitions, and to the study of bottlenecks and deficiencies in organizational structure. For instance, the discussion of the proper ratio of developers to testers, which is moot for teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.

Third, the observation that media use varies depending on the kinds of expectations placed on the user complements and helps explain the earlier finding that electronic repositories are incomplete and often misleading \cite{Aranda2009}. If researchers focus on a single electronic repository, they are likely to miss systematically important areas of organizational coordination: they will be blind to some people and expectations. Depending on the research questions they address, this may be a serious concern. Judging by our two cases, mining issue tracking systems should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate, and focus only on those who emit messages in the system.

\emph{Practice:} Our qualitative data confirm, in the software development domain, the old sociological finding that the role specialization trend as an organization grows is hard to fight. This is not necessarily bad, but teams of would-be generalists (which Agile techniques often propose) need to calculate the long-term impacts of seemingly necessary short-term specializations. A role that makes sense in the present to deal with a dysfunction may spin off to become a compartmentalized department later on, with its own political, budgetary, and organizational goals. This may not be in the interest of the current members of the organization.

Additionally, we found that many practitioners do not have an awareness of the full scope of their peers' activities, and that this lack of awareness leads them to misunderstand the benefits that they bring to the organization. It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that their formal position does not match the expectations placed on them. In these cases, a reductionism to position labels is harmful.

@@

The two organizations we studied were small (less than fifty members), geographically co-located for the most part, and North American. Therefore, the generalizability of our findings might be questioned. However, most of our findings are supported by a theoretical framework that is not built upon these particular characteristics, but is instead general and interdisciplinary. We expect that some of the particular expectation clusters will vary in other groups, depending on their contextual factors, but we also expect the underlying rationale and observations to hold across them.

@@

Throughout this paper, we defined role as a ``set of expectations.'' However, an ``expectation'' is a subjective concept: we expect many things from many people, and an objective articulation of these expectations may never be quite right, missing some attributes or formulating an expectation at the wrong level of detail. We acknowledge this threat, but we see it as a necessary evil in dealing with this subject. We hope that with more research in the area, a consensus will emerge on the right level of analysis for expectations in software development work.

>>>>

There was no simple role classification in the organizations we studied. However, it appears that some patterns in role definitions would emerge with further studies. We can see the beginnings of such patterns in Table \ref{tableExpectations}, which shows the expectations placed on the people we interviewed. The columns in the table represent our interviewees and the rows represent common or significant expectations. The reader should note that the list of expectations is necessarily incomplete for the general case. It is also somewhat subjective, as expectations were not always clearly articulated, and our participants did not agree on their expectation assignments in every case.

One pattern concerns the technical core of the organizations' products. Work on this area was reserved almost exclusively for developers and their technical and team leads. However, the expectations placed on leads were more spread out: technical leads focused more on architecture and design issues, whereas team leads emphasized coordination activities. There were also purely organizational roles that do not need any domain or technical expertise. But in both organizations some key people had more than one kind of expertise (technical, domain, and organizational), and these people seemed vital for the functioning of the organization as they transitioned between boundaries with ease. A representative example is a Gallium Field Specialist (the second column in the Support \& Field cluster), who straddled all three kinds of expertise. Several people in different areas pointed out his relevance; in the words of one informant, ``the company depends on [him] so much he should have armed guards behind him taking him to and from work.''

@@

\emph{Product Management at Gallium:} There were three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead. The latter documented the requirements of new functionality, provided a historical perspective on design decisions, and helped shape the organizational structure. Of the two Product Managers, one had an inside-looking approach, whereas the other one had an outside-looking approach: he had a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimized the organization. He also understood how the users would interact with the product and helped design the user interface to bring it closer to the users' mental model. Other people in the organization also performed some Product Management activities unofficially: Customer Services Representatives and Field Specialists often persuade developers to focus on the features that cause them the most pain, for example. In sum, although the three employees above have similar goals, the expectations placed on them vary considerably, are not interchangeable, and are not their exclusive prerogative.

\emph{Team Leading at Gallium and Cobalt:} Every team in the sites we studied had a Team Lead, but this leadership role was tinged by the characteristics of the team in question. In Cobalt, team membership is fluid---every developer may be expected to work on any project, according to need---and some Team Leads performed coordination tasks, others had a dormant leadership position, waiting for new issues to arise, and yet others were more like ``first responders'' than coordinators, requesting assistance only if their work load became excessive. In Gallium, there were three Team Leads at the time of our observations. Their roles were quite different. The Team Lead of the Alpha Team would spend about half of his time developing software, and the other half in meetings, performing estimation tasks, and analyzing the performance of his team. A second Team Lead was in charge of a product about to be discontinued, but that responsibility took a small part of his time. He was mostly associated with the Beta Team, but since that team had an abundance of ``leads'' (it included the two official Tech Leads of the organization), his coordination and reporting expectations were minimal. The third Team Lead had one other team member, a newcomer to the organization, and his leadership expectations consisted mostly of mentoring and training rather than coordination and monitoring.

\emph{Developers at Gallium and Cobalt:} The quintessential role in the software development literature was one of the hardest to pin down. A developer, of course, is expected to write code, but what else? There is a multitude of answers to that question: some developers are expected to perform code reviews, some to provide architectural input, some to carry their own requirements analysis, some to socialize, some to interface with customers. None of the developers in our sample were expected to perform all of the above.

\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization. But the stuff that falls through the cracks was different in both sites, and therefore, the expectations were different. In Cobalt, much of this referred to sheltering the organization from bureaucratic processes and constraints. In Gallium, it partly meant helping create and shape new and needed groups, and partly to perform activities that nobody else was tasked with doing, such as some kinds of documentation.

\emph{Project Management at Gallium:} These Project Managers were a counterexample to our findings, to some extent. Although there were distinctions between them (one had a Project Management Leader role, another created templates and instituted new processes, and all, including a third and more junior Project Manager, implemented them), the similarities in their roles seemed to outweigh the differences. The main reason seemed to be that they brought a fairly specific and standardized way of doing things with them, in the form of a Project Management Body of Knowledge and certification. In other words, the institutionalization of their role was strong. They also strived to define their role with precision, and to enforce their definition in the rest of the organization. No other role that we examined had the emphasis on role definition and consistency that the Gallium Project Managers had.

@@

We found that expectations are grouped in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors, and that these factors interplay and provide feedback to each other.

\emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it helps shape role definitions. We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital. It quickly expanded and distributed its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them with newcomers. Before this influx, the division of labour was flexible and organic: most organization members did a bit of everything. After the influx, role definitions shifted towards a more stereotypical differentiation.

\emph{Historical factors:} Previous events in the history of the organization shape its current grouping of expectations, in the same way that alterations early in the life of other organic systems shape their later structure. Some developers that have been part of the organization for a long time are not just developers, they are also mentors, historians, and political players, despite what the organizational chart says about them. This affects the expectations placed on them. During the Gallium downsizing, some people were demoted to keep the company afloat. At some point, one of them had been the head of the QA department (which was dispersed) and, later, the Program Manager. At the time of our observations he was a QA Analyst, yet he was still expected to be involved in strategic activities that did not match the institutional view of what a QA Analyst is supposed to do.

\emph{Structural factors:} A person's position in the organizational structure gives her a certain power and awareness that others may not have. Her structural position leads others to expect from her certain behaviours: to play a bridging, directing, or even confrontational role. Many of the expectations placed on the head of the Operations group at Cobalt spawned from his structural position. He was expected to provide news of the operational status of Cobalt's services, to deal with and shelter the Software group from system administration concerns, and to control the gateway that determines whether a product is ready to be released or not---that is, to have final say on the operational quality of the product. These expectations come from his position in the organization. It would be hard for someone in a different position to satisfy them.

\emph{Personal factors:} A person's background, preferences, and personality shapes the expectations she receives from others. Perhaps her background is deficient for the fulfillment of her initial role, and others adapt by expecting different or lesser things from her. Alternatively, her traits (curiosity, proactivity, a talent) may lead her to take on responsibilities that would not otherwise be placed on her given institutional, historical, or structural factors. For example, the Project Administrator at Cobalt was hired to perform clerical tasks, but his own initiative led him to gradually take more and greater responsibilities. The organization learned to be dependent on such behaviour. When he temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement. Projects suffered quickly and significantly as the replacement only focused on the expectations that were officially placed on him. We should also note that at the time of our observations, the Project Administrator position was undergoing an official redefinition in order to match real expectations. In this way, personal factors were leading to historical and structural changes, so that future enactors of the role played by the Project Administrator would be affected by them.

@@

However, a person's position has limited power in defining their role. As we have seen above, the expectations placed on a person also come from historical and personal factors. An example from our observations is one of the Gallium Product Managers. He was hired to steer and coordinate the development of the company's products, but he had no experience in software development. He thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization, he struggled to understand the level of control he was expected to have. Over time, he adjusted his behaviour, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of others with similar positions.

@@

Regarding organizational growth, it has been recognized as a factor in specialization for a long time, and our data are far from surprising. In fact, it should be clear from pure argument that growth drives role evolution. The smallest organization is formed by a single person, and that person should perform all the activities required for the functioning of the organization, from sales to debugging. As the organization obtains more members, labour begins to be divided, so that upon reaching even a handful of members, it is rare for a software group to perform undifferentiated labour. Furthermore, growth carries a coordination weight: larger organizations need to assign some of their members the role of coordinating the work of others, or of interfacing between groups. That is, larger organizations have needs that are caused by their size. With organizations that are similar to those we observed, some degree of specialization is well underway. We expect that, should we perform this study in a larger organization, we would discover further role specialization.

Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest. In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation. The silos that the group wanted to break were project-based. They discovered that assigning one person per project introduced vulnerabilities (when something went wrong in a project, only one person could intervene) and rework (as many projects consisted of similar modules that had to be developed independently). To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem. The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate.

The other important factor driving role evolution was a mismatch between expectations and behaviour. The fact that expectations are placed does not mean that they are satisfied. There are many potential reasons for their lack of satisfaction: for instance, the person upon whom they are placed may not know or understand the expectation, it could be beyond her means, or it could be in conflict with other expectations. Therefore, an assignment of expectations that works in theory may not work in practice. Similarly, an assignment that does \emph{not} work in theory may work in practice, due to the initiative or motivation of people who contribute beyond their expectations. However, in both cases, with time, organization members grow to expect the typical performance of others, and a new expectation (and hence, a new role) takes hold.

Several examples from our data above illustrate this phenomenon and we offer one more here. The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being a contact for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction. In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization. As a result, he now plays an essential role in the development of the group's products and in the maintenance of its relations with its users, a role that the rest of the organization depends on for its good functioning.

@@

We expected to find that people enacting similar roles use similar communication media, and that they communicate with people enacting interfacing roles through other media. Although we found support for this claim, a biasing factor was the prevalence of face-to-face interactions in both of our field sites. Regardless of role, people communicated face-to-face regularly. Stand-up meetings, walks across the hall, and group meetings were commonplace.

Regarding other media, we found role-based differences in medium selection and use. The broadest medium---the one that had the greatest use among the most kinds of roles---was the issue tracking system. People with expectations regarding code production, quality assurance, customer service, field implementation, product management, and performance monitoring all used the issue tracking system to some extent. Its use was not uniform: some group members used it to emit messages, others were mostly listeners, or they were more interested in the aggregate activity of the team. This finding mirrors the observation of De Souza \emph{et al.}\ regarding issue tracking systems as boundary objects \cite{deSouza2003}.

The use of other media was more localized. In Gallium, IRC use was common among some developer teams: each team had one channel, and these were not frequented by any non-developers. Customer Support staff used Google Documents to keep track of its cases, and to provide an ongoing documentation of recent events and requests for support. Email and phone were more heavily used in the management side of the organization: developers and QA analysts rarely made or received phone calls, and they obtained most of the information they needed about their projects through the issue tracking system.

@@

We also explored role evolution and found that it is driven by organizational growth, and by mismatches between expectations and behaviour. Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites across roles.

@@

\emph{Research:} First, the fact that role definitions are not institutionally uniform and respond to local (historical, structural, personal) factors should help clarify common misunderstandings and assumptions made in research. A stereotypical developer, team lead, or tester, is rarely found in practice. Instead, thinking in terms of tasks and expectation groupings will help us ground our studies to software development in the real world.

Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation pairings (expectations that make good sense, organizationally, to be placed on the same individual). This in turn should lead to better role definitions, and to the study of bottlenecks and deficiencies in organizational structure. For instance, the discussion of the proper ratio of developers to testers, which is moot for teams where there is no QA and developers do all the testing, can be explored in terms of the proper ratio of people expected to write code to people expected to write integration tests or to determine the level of quality of the product.

Third, the observation that media use varies depending on the kinds of expectations placed on the user complements the earlier finding that repositories are incomplete and often misleading \cite{Aranda2009}. If researchers focus on a single repository, they are likely to systematically miss important areas of coordination: they will be blind to some people and expectations. Depending on their research questions, this may be a serious concern. Judging by our two cases, mining issue tracking systems should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate activity.

\emph{Practice:} Our data confirm, in the software development domain, the old sociological finding that the role specialization trend as an organization grows is hard to fight. This is not necessarily bad, but teams of would-be generalists (which Agile techniques often propose) need to calculate the long-term impacts of seemingly necessary short-term specializations. A role that makes sense in the present to deal with a dysfunction may spin off to become a compartmentalized department later on, with its own political, budgetary, and organizational goals. This may not be in the interest of the current members of the organization.

Additionally, we found that practitioners may not have an awareness of the full scope of their peers' activities, and that this lack of awareness leads them to misunderstand the benefits that they bring to the organization if they make judgments based on labels. It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that their formal position does not match the expectations placed on them. In these cases, a reductionism to position labels is harmful.

@@

The two organizations we studied were small (less than fifty members), geographically co-located for the most part, and North American. Therefore, the generalizability of our findings might be questioned. However, most of our findings are supported by a theoretical framework that is not built upon these particular characteristics, but is instead general and interdisciplinary. We expect that some of the particular expectation pairings will vary in other groups, but we also expect the underlying rationale and observations to hold across them.

@@

Throughout this paper, we defined role as a ``set of expectations.'' However, an ``expectation'' is a subjective concept: we expect many things from many people, and an objective articulation of these expectations may never be quite right, missing some attributes or formulating an expectation at the wrong level of detail. We acknowledge this threat, but it is a necessary evil. We hope that with more research in the area, a consensus will emerge on the right level of analysis for expectations in software organizations.
